Unshuffling Permutations

### 2015-09-15 article.tex
Ajouts dans conclusion et dans introduction

\begin{abstract}
  A permutation is said to be a square if it is the union of two order-isomorphic
  patterns.
  The definition is intended to be the natural counterpart to the ordinary shuffle
  of words and languages.
  In this paper, we tackle the problem of recognising square permutations from both
  the point of view of algebra and algorithmic.
  On the one hand, we present some algebraic properties of the shuffle product of
  permutations and exhibit a bijection between square $(213,231)$-avoiding permutations
  and square binary words.
  On the other hand, we prove that recognising square permutations is \NPC.
\end{abstract}

@@

Given three words $u$, $v$ and $w$, $u$ is said to a
\emph{shuffle} of $v$ and $w$ if it
can be formed by interleaving the letters from $v$ and $w$ in a way
that maintains the left-to-right ordering of the letters from each word
(it is a classical textbook exercice to
design an efficient dynamic programming algorithm that determines
whether $u$ is a shuffle of $v$ and $w$).
A word $u$ is said to be a \emph{square} (w.r.t. the shuffle) if there exists
a word $v$ such that $u$ is a shuffle of $v$ with itself.

This paper is intended to study a natural generalisation of the shuffle
to permutations.
Given permutations $\pi$, $\mu$ and $\nu$, $\pi$ is said to be a shuffle of
$\mu$ and $\nu$ if $\pi$ (viewed as a word) is a shuffle of two words that
are order-isomorphic to $\mu$ and $\nu$.
For example, $\pi$ is the shuffle
of $\mu=$ and $\nu=$ as
$\pi$.
A permutation $\pi$ is said to be a square (w.r.t. the shuffle) if there exists
a permutation $\mu$ such that $\pi$ is a shuffle of $\mu$ with itself.
For example, $\pi = 241356$ is a square as it is the shuffle of $213$ with itself
(\emph{e.g.},
$\pi = \dot{2}\ddot{4}\dot{1}\ddot{3}\ddot{5}\dot{6}$, where
$\dot{i}$ and $\ddot{i}$),
whereas $\pi = 4123$ is not.

Formally, the shuffle of two words $u$ and $v$ of $A^*$ is the language
$u \shuffle v$ consisting of all words
$u_1 v_1 u_2 v_2 \ldots u_k v_k$, where $k \geq 0$ and
the $u_i$ and the $v_i$ are the words  of $A^*$
such that $u = u_1 u_2 \ldots u_k$ and
$v = v_1 v_2 \ldots v_k$.
We first briefly review the key results for words
that arise in our context.
Given words $u$, $v_1$ and $v_2$, it can be tested in
$O\left(|u|^2 / \log(|u|)\right)$ time whether or not
$u \in v_1 \shuffle v_2$~\cite{Leeuwen:Nivat:IPL:1982}.
(To the best of our knowledge, the first
$O(|u|^2)$ time algorithm appeared in~\cite{Mansfield:DAM:1983}.
This algorithm can easily be extended to check in polynomial-time
whether or not a word is the shuffle of any fixed number of given words.)
The shuffle $u \shuffle v$ of words $u$ and $v$ can be computed in
$O\left((|u|+|v|) \; \binom{|u|+|v|}{|u|}\right)$
time~\cite{Spehner:TCS:1986}.
An improvement and generalization has been proposed in
\cite{Allauzen:IGM:2000}, where it is proved that,
given words $u_1, u_2, \ldots, u_k$, the shuffle
$\shuffle_{i=1}^{k} u_i$ can be computed in
$O\left(\binom{|u_1|+|u_2|+\ldots+|u_k|}{|u_1|,|u_2|,\ldots,|u_k|}\right)$ time.
Given words $u, v_1, v_2, \ldots, v_n \in A^*$, it is however \NPC to
decide whether or not
$u \in \shuffle_{i=1}^{k} v_i$.
\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}.
This remains true even if the alphabet has size $3$~\cite{Warmuth:Haussler:JCSS:1984}.
Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that this problem remains
\NPC even if all
words $v_1, v_2, \ldots, v_n$ are identical,
thereby proving that,
for two words $u$ and $v$, it is \NPC to decide whether or not $u$
is in the iterated shuffle of $v$.
Again, this remains true even if the alphabet has size $3$.
As for squares, given a word $z$ it is \NPC to determine whether
there exists a word $x$ such that $z$ is the shuffle of $x$ with itself
\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}.

$123$-avoiding (resp. $321$-avoiding) permutations are the union of two decreasing
(resp. incresing) patterns.

@@

There are a number of further directions of investigation in this general subject.
We mention several -~not necessarily new~- open problems that are, in our opinion,
the most interesting.
How many permutations of $S_{2n}$ are squares?
How many $(213,231)$-avoiding permutations of $S_{2n}$ are squares?
(Equivalentely, how many binary strings of length $2n$ are squares;
see Problem~4 in \cite{Henshall:Rampersad:Shallit:2011})?
How hard is the problem of deciding whether a $(213,231)$-avoiding permutation
is a square (see Problem~4 in \cite{Henshall:Rampersad:Shallit:2011})?)
Given two permutations $\pi$ and $\sigma$, how hard is the problem of deciding
whether $\pi$ is the union of two patterns order-isomorphic to $\sigma$?

>>>>

\begin{abstract}
    A permutation is said to be a square if it is the union of two
    order-isomorphic patterns. The definition is intended to be the
    natural counterpart to the ordinary shuffle of words and languages.
    In this paper, we tackle the problem of recognizing square
    permutations from both the point of view of algebra and algorithmic.
    On the one hand, we present some algebraic properties of the shuffle
    product of permutations and exhibit a bijection between square
    $(213,231)$-avoiding permutations and square binary words. On the
    other hand, we prove that recognizing square permutations is \NPC.
\end{abstract}

@@

Given three words $u$, $v$ and $w$, $u$ is said to a \emph{shuffle} of
$v$ and $w$ if it can be formed by interleaving the letters from $v$ and
$w$ in a way that maintains the left-to-right ordering of the letters
from each word (it is a classical textbook exercise to design an
efficient dynamic programming algorithm that determines whether $u$ is a
shuffle of $v$ and $w$). A word $u$ is said to be a \emph{square} (w.r.t.
the shuffle) if there exists a word $v$ such that $u$ is a shuffle of $v$
with itself.
\smallskip

This paper is intended to study a natural generalization of the shuffle
to permutations. Given permutations $\pi$, $\mu$ and $\nu$, $\pi$ is
said to be a shuffle of $\mu$ and $\nu$ if $\pi$ (viewed as a word) is
a shuffle of two words that are order-isomorphic to $\mu$ and $\nu$. For
example, $\pi$ is the shuffle of $\mu=$ and $\nu=$ as $\pi$. A
permutation $\pi$ is said to be a square (w.r.t. the shuffle) if there
exists a permutation $\mu$ such that $\pi$ is a shuffle of $\mu$ with
itself. For example, $\pi = 241356$ is a square as it is the shuffle of
$213$ with itself (\emph{e.g.},
$\pi = \dot{2}\ddot{4}\dot{1}\ddot{3}\ddot{5}\dot{6}$, where $\dot{i}$
and $\ddot{i}$), whereas $\pi = 4123$ is not.
The shuffle product defined and considered in this paper is different
from other existing and well-known shuffle products on permutations.
For instance, in~\cite{DHT:IJAC:2002}, the authors defined the
{\em shifted shuffle product}. In this product $\pi$ is a shuffle of
$\mu$ and $\nu$ if $\pi$ is the shuffle, as words, of $\mu$ and the
word obtained by incrementing all letters of $\nu$ by the size of $\mu$.
It is a simple exercise to prove that, give a permutation $\pi$, the
decision problem to test if $\pi$ is a square for the shifted shuffle
product is linear.
\smallskip

Formally, the shuffle of two words $u$ and $v$ of $A^*$ is the language
$u \shuffle v$ consisting of all words $u_1 v_1 u_2 v_2 \ldots u_k v_k$,
where $k \geq 0$ and the $u_i$ and the $v_i$ are the words  of $A^*$
such that $u = u_1 u_2 \ldots u_k$ and $v = v_1 v_2 \ldots v_k$. We
first briefly review the key results for words that arise in our context.
Given words $u$, $v_1$ and $v_2$, it can be tested in
$O\left(|u|^2 / \log(|u|)\right)$ time whether or not
$u \in v_1 \shuffle v_2$~\cite{Leeuwen:Nivat:IPL:1982}.
(To the best of our knowledge, the first $O(|u|^2)$ time algorithm a
ppeared in~\cite{Mansfield:DAM:1983}. This algorithm can easily be
extended to check in polynomial-time whether or not a word is the
shuffle of any fixed number of given words.) The shuffle $u \shuffle v$
of words $u$ and $v$ can be computed in
$O\left((|u|+|v|) \; \binom{|u|+|v|}{|u|}\right)$ time~\cite{Spehner:TCS:1986}.
An improvement and generalization has been proposed in
\cite{Allauzen:IGM:2000}, where it is proved that, given words
$u_1, u_2, \ldots, u_k$, the shuffle $\shuffle_{i=1}^{k} u_i$ can be
computed in
$O\left(\binom{|u_1|+|u_2|+\ldots+|u_k|}{|u_1|,|u_2|,\ldots,|u_k|}\right)$
time. Given words $u, v_1, v_2, \ldots, v_n \in A^*$, it is however \NPC
to decide whether or not $u \in \shuffle_{i=1}^{k} v_i$.
\cite{Mansfield:DAM:1983,Warmuth:Haussler:JCSS:1984}.
This remains true even if the alphabet has size $3$~\cite{Warmuth:Haussler:JCSS:1984}.
Of particular interest,
it is shown in~\cite{Warmuth:Haussler:JCSS:1984} that this problem remains
\NPC even if all
words $v_1, v_2, \ldots, v_n$ are identical,
thereby proving that,
for two words $u$ and $v$, it is \NPC to decide whether or not $u$
is in the iterated shuffle of $v$.
Again, this remains true even if the alphabet has size $3$.
As for squares, given a word $z$ it is \NPC to determine whether
there exists a word $x$ such that $z$ is the shuffle of $x$ with itself
\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}.
\smallskip

$123$-avoiding (resp. $321$-avoiding) permutations are the union of two decreasing
(resp. increasing) patterns.

@@

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, how many binary
strings of length $2n$ are squares; see Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(see Problem~4 in \cite{Henshall:Rampersad:Shallit:2011})?) Given two
permutations $\pi$ and $\sigma$, how hard is the problem of deciding
whether $\pi$ is the union of two patterns order-isomorphic to $\sigma$?
\smallskip

Besides, in this work, we investigate some algebraic properties of the
shuffle of permutations which have direct combinatorial consequences.
However, this work raises several additional classical algebraic
questions. Among these, one can ask for a complete algebraic study of
$\QQ[S]$ as a graded associative algebra for the shuffle  product
$\SHUFFLE$. Describing a generating family for $\QQ[S]$, defining
multiplicative bases of $\QQ[S]$, and determining whether $\QQ[S]$ is
free as an associative algebra are worthwhile questions.
